Effectful Semantics for Substructural SSA

### 2023-11-16 pldi24/paper.tex
Added optimisation text

\begin{itemize}
\item We give a type system for SSA-style programs, which has a full
substitution theorem and equational theory. This language includes substructural
types, both for controlling effects and tracking ownership in values, and our
strong substitution principle makes formulating inlining and rewriting
optimisations easier than in the traditional CFG-based formulations. 
\item We give an axiomatic semantics for our language, by showing that it has a
compositional intepretation in any effectful category with an Elgot structure.
We use this structure to give generic proofs of the soundness of substitution
and rewriting properties.
\item Our abstract semantics justifies a variety of powerful optimisation
techniques, including E-graph rewriting, inlining, loop fusion, hoisting, and
strength reduction. 
\item We show how Elgot monads give rise to instances of our categorical
semantics. We use this to derive semantics from a general trace monad, and show
that various monad transformers such as state transformers preserve the needed
Elgot structure. 
\item We use these tools to demonstrate the existence of a variety of concrete
models satisfying these categorical axioms. Starting first from simple languges
such as state plus printing, we also show how more challenging semantics such as
the TSO semantics for weak memory also fit into our framework. This thus gives
rise to an SSA-based IR with support for weak memory operations, which is fully
semantically-justified.
\item Our weak memory semantics and our results about Elgot monads are
formalised in the Lean theorem prover.
\end{itemize}

@@

Now that we've given a semantics to \isotopessa programs, we can begin to use it
to develop an equational theory for performing optimizations. Since we're
approaching things from a type-theoretic perspective, we begin by addressing the
semantic versions of the standard syntactic metatheorems of weakening and
substitution. In particular, semantic weakening can be stated in the expected
manner: if we have an expression or block which is well typed in both \(\Gamma\)
and \(\Delta\), where \(\cwk{\Gamma}{\Delta}\), then evaluating it in \(\Gamma\)
is the same as first strengthening \(\Gamma\) to \(\Delta\), dropping unused
affine variables, and then evaluating it in \(\Delta\).
\begin{lemma}[Semantic Weakening] 
  If \(\cwk{\Gamma}{\Delta}\), then 
  \begin{itemize}
    \item For all \(\hasty{\Delta}{p}{a}{A}{q}\),
    \[
      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
      ; \dnt{\hasty{\Delta}{p}{a}{A}{q}}
      = \dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
    \]
    \item For all \(\haslb{\Delta}{t}{\ms{L}}\),
    \[
      \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
      ; \dnt{\haslb{\Delta}{t}{\ms{L}}}
      = \dnt{\haslb{\Gamma}{t}{\ms{L}}}  
    \]
  \end{itemize}

\TODO{some text about upgrade}

@@

\subsection{Optimizations}

\TODO{this:}

\TODO{substitution \(\implies\): loop hoisting, global value numbering, some dead code elimination}

\TODO{rewriting \(\implies\) strength reduction, E-graph optimization, see \cite{cranelift}}

\TODO{inlining}

\TODO{loop fusion}

\TODO{unreachable code elimination}

\TODO{more dead code elimination}

>>>>

\begin{itemize}
\item We give a type system for SSA-style programs, which has a full
substitution theorem and equational theory. This language includes substructural
types, both for controlling effects and tracking ownership in values, and our
strong substitution principle makes formulating inlining and rewriting
optimisations easier than in the traditional CFG-based formulations. 
\item We give an axiomatic semantics for our language, by showing that it has a
compositional intepretation in any effectful category with an Elgot structure.
We use this structure to give generic proofs of the soundness of substitution
and rewriting properties.
\item Our abstract semantics justifies a variety of powerful optimisation
techniques, inlining, loop fusion, hoisting, and strength reduction. 
\item We show how Elgot monads give rise to instances of our categorical
semantics. We use this to derive semantics from a general trace monad, and show
that various monad transformers such as state transformers preserve the needed
Elgot structure. 
\item We use these tools to demonstrate the existence of a variety of concrete
models satisfying these categorical axioms. Starting first from simple languges
such as state plus printing, we also show how more challenging semantics such as
the TSO semantics for weak memory also fit into our framework. This thus gives
rise to an SSA-based IR with support for weak memory operations, which is fully
semantically-justified.
\item Our weak memory semantics and our results about Elgot monads are
formalised in the Lean theorem prover.
\end{itemize}

@@

Now that we've given a semantics to \isotopessa programs, we can begin to use it
to develop an equational theory for performing optimizations. Since we're
approaching things from a type-theoretic perspective, we begin by addressing the
semantic versions of the standard syntactic metatheorems of weakening and
substitution.

\todo{Move weakening and upgrade lemmas to appendix}

% In particular, semantic weakening can be stated in the expected
% manner: if we have an expression or block which is well typed in both \(\Gamma\)
% and \(\Delta\), where \(\cwk{\Gamma}{\Delta}\), then evaluating it in \(\Gamma\)
% is the same as first strengthening \(\Gamma\) to \(\Delta\), dropping unused
% affine variables, and then evaluating it in \(\Delta\).
% \begin{lemma}[Semantic Weakening] 
%   If \(\cwk{\Gamma}{\Delta}\), then 
%   \begin{itemize}
%     \item For all \(\hasty{\Delta}{p}{a}{A}{q}\),
%     \[
%       \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
%       ; \dnt{\hasty{\Delta}{p}{a}{A}{q}}
%       = \dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
%     \]
%     \item For all \(\haslb{\Delta}{t}{\ms{L}}\),
%     \[
%       \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
%       ; \dnt{\haslb{\Delta}{t}{\ms{L}}}
%       = \dnt{\haslb{\Gamma}{t}{\ms{L}}}  
%     \]
%   \end{itemize}
% 
%   Similarly, if \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then
%   \[
%     \dnt{\haslb{\Gamma}{t}{\ms{L}}}
%     ; \dnt{\lwk{\ms{L}}{\ms{K}}}
%     = \dnt{\haslb{\Gamma}{t}{\ms{K}}}
%   \]
% \end{lemma}
% 
% \TODO{some text about upgrade}
% 
% \begin{lemma}[Semantic Upgrade]
%   For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
%   \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\]
% \end{lemma}

@@

\subsection{Optimisations}

Because we have a denotational semantics, it is easy to validate
equations between terms.

Even the basic proof of the soundness of substitution already lets us
prove the soundness of many fundamental compiler optimisations, such
as global value numbering (let-introduction for pure expressions),
some forms of dead code elimination (i.e., eliminating unused affine
bindings), and strength reduction with loop hoisting (let-introduction
for relevant expressions).

Substitution only applies to central terms, because those are the only
terms for which substitution is sound. Furthermore, there are many
other optimisations that follow from the beta-eta theory of the
language, as well as specific properties of model-dependent operations
(such as memory operations). In the appendix, we prove a general
rewriting theorem which lifts equations between terms into a full
congruence. This lets us justify optimisations such as scalar
replacement of aggregates, eliminating redundant writes, reordering
independent memory operations, and memory-to-register transformations.

Because we have a type-theoretic presentation of control-flow graphs,
we also have a substitution principle and equational theory for
them. This lets us show the soundness of inlining, loop unrolling,
loop fusion, dead-code elimination, and other optimisations which
change the structure of the control-flow graph.

Note that we have not yet proved anything about the dataflow analyses
used to find the rewrites -- we have only shown that the rewrites
themselves are semantically justified.
